Dewlap color variation in \textit{Anolis sagrei} is maintained among habitats within islands of the West Indies

### 2020-08-08 ms/main.tex
jonathans comments implemented

\title{Dewlap color variation in \textit{Anolis sagrei} is maintained between habitats within islands of the West Indies}
\author{
    \textsc{Rapha\"{e}l Scherrer$^{1,3}$ \thanks{Corresponding author: r.scherrer@rug.nl}, Colin M. Donihue$^{1,4}$, }\\
	\textsc{R. Graham Reynolds$^2$, Jonathan B. Losos$^{1,4}$ and Anthony J. Geneva$^{1,5}$} \\[1ex]
	%\thanks{A thank you} \\[1ex] % Your name
	\normalsize $^1$ Department of Organismic and Evolutionary Biology and Museum of Comparative Zoology \\ \normalsize Harvard University, Cambridge, MA, USA \\
	\normalsize $^2$ Department of Biology, University of North Carolina Asheville, Asheville, NC, USA\\ 
	\normalsize $^3$ Current address: Groningen Institute for Evolutionary Life Sciences,\\
	\normalsize Groningen, The Netherlands\\
	\normalsize $^4$ Current address: Department of Biology, Washington University, St. Louis, MO, USA\\
	\normalsize $^5$ Current address: Department of Biology, Center for Computational and Integrative Biology,\\ \normalsize Rutgers University--Camden, Camden, NJ, USA
}
\date{} % Leave empty to omit a date

@@

\textbf{Keywords} --- \textit{Anolis}, reflectance, local adaptation, sexual signal, supervised machine learning

>>>>

\title{Dewlap color variation in \textit{Anolis sagrei} is maintained among habitats within islands of the West Indies}
\author{
    \textsc{Rapha\"{e}l Scherrer$^{1,3}$ \thanks{Corresponding author: r.scherrer@rug.nl}, Colin M. Donihue$^{1,4}$, }\\
	\textsc{R. Graham Reynolds$^2$, Jonathan B. Losos$^{1,4}$ and Anthony J. Geneva$^{1,5}$} \\[1ex]
	%\thanks{A thank you} \\[1ex] % Your name
	\normalsize $^1$ Department of Organismic and Evolutionary Biology and Museum of Comparative Zoology \\ \normalsize Harvard University, Cambridge, MA, USA \\
	\normalsize $^2$ Department of Biology, University of North Carolina Asheville, Asheville, NC, USA\\ 
	\normalsize $^3$ Current address: Groningen Institute for Evolutionary Life Sciences,\\
	\normalsize University of Groningen, Groningen, The Netherlands\\
	\normalsize $^4$ Current address: Department of Biology, Washington University, St. Louis, MO, USA\\
	\normalsize $^5$ Current address: Department of Biology, Center for Computational and Integrative Biology,\\ \normalsize Rutgers University--Camden, Camden, NJ, USA
}
\date{} % Leave empty to omit a date

@@

\textbf{Keywords} --- reflectance, adaptation, sexual signal, machine learning, polymorphism

### 2020-08-08 ms/sections/abstract.tex
jonathans comments implemented

Animal signals evolve in an ecological context. Moreover, locally adapting animal sexual signals can be especially important for initiating or reinforcing reproductive isolation during the early stages of speciation. Dewlap color in \textit{Anolis} lizards can be highly variable between populations in relation to both biotic and abiotic adaptive drivers, albeit at relatively large geographical scales. At smaller scales, gene flow is likely to erase phenotypic differentiation between localities. Here, we investigated local adaptation of the dewlap across habitat-types at a small spatial scale, as this may give an indication of how conditions for the early stages of speciation may be met. We explored variation in dewlap coloration in the most widespread species of anole, \textit{Anolis sagrei}, across three characteristic habitats spanning the Bahamas and the Cayman Islands. Using reflectance spectrometry as well as supervised machine learning, we found significant differences in spectral properties of the dewlap between habitats within small islands, sometimes over very short distances. Passive divergence in dewlap phenotype associated with isolation-by-distance did not explain our results. On the other hand, these habitat-specific dewlap differences varied in magnitude and direction across islands, and thus our primary test for adaptation -- parallel responses across islands -- was falsified. We suggest, however, that selection could be involved and present several ways this might have happened, including sexual selection. (This one may be too much of a leap) While at present, populations from these different habitats probably experience too much gene flow to follow distinct evolutionary lineages, should additional barriers arise between habitat-specific populations, the observed disruptive selection on dewlap coloration may facilitate speciation.

>>>>

Animal signals evolve in an ecological context. Moreover, locally adapting animal sexual signals can be especially important for initiating or reinforcing reproductive isolation during the early stages of speciation. Dewlap color in \textit{Anolis} lizards can be highly variable between populations in relation to both biotic and abiotic adaptive drivers, albeit at relatively large geographical scales. At smaller scales, gene flow is likely to erase phenotypic differentiation between localities. Here, we investigated differentiation of dewlap coloration among habitat-types at a small spatial scale, within multiple islands of the West Indies, as this may give new insights into the local scale at which adaptation is possible. We explored variation in dewlap coloration in the most widespread species of anole, \textit{Anolis sagrei}, across three characteristic habitats spanning the Bahamas and the Cayman Islands. Using reflectance spectrometry as well as supervised machine learning, we found significant differences in spectral properties of the dewlap between habitats within small islands, sometimes over very short distances. Passive divergence in dewlap phenotype associated with isolation-by-distance did not explain our results. On the other hand, these habitat-specific dewlap differences varied in magnitude and direction across islands, and thus our primary test for adaptation -- parallel responses across islands -- was falsified. We suggest, however, that selection could be involved in several ways, including sexual selection. Our results shed new light on the scale at which signal color polymorphism can be maintained in the presence of gene flow, and the relative role of local adaptation and other processes in driving these patterns.

### 2020-08-08 ms/sections/acknowledgments.tex
jonathans comments implemented

Collection permission was granted by the Bahamas Environment, Science and Technology Commission, the Bahamas National Trust, and the Cayman Islands Department of the Environment. The authors thank Sofia Prado-Irwin, Pavitra Muralidhar, Nicholas Herrmann, Richard E. Glor, Alberto R. Puente-Rol\'{o}n, Kevin Aviles-Rodriguez, Kristin Winchell, Jason Fredette and Melissa Kemp for assistance in the field and Max Lambert and James Stroud for helpful discussions. Funding for this work was provided by the Templeton Foundation (to JBL), NSF DEB~\#1927194 (to JBL and AJG), NSF DEB~\#1500761 (to AJG), NSF DBI~\#1609284 (to CMD), and a Harvard Museum of Comparative Zoology Putnam Expedition Grant (to RGR).

>>>>

Collection permission was granted by the Bahamas Environment, Science and Technology Commission, the Bahamas National Trust, and the Cayman Islands Department of the Environment. The authors thank Sofia Prado-Irwin, Pavitra Muralidhar, Nicholas Herrmann, Richard E. Glor, Alberto R. Puente-Rol\'{o}n, Kevin Aviles-Rodriguez, Kristin Winchell, Jason Fredette and Melissa Kemp for assistance in the field and Pratik Gupte, Max Lambert and James Stroud for helpful discussions. Funding for this work was provided by the Templeton Foundation (to JBL), NSF DEB~\#1927194 (to JBL and AJG), NSF DEB~\#1500761 (to AJG), NSF DBI~\#1609284 (to CMD), and a Harvard Museum of Comparative Zoology Putnam Expedition Grant (to RGR).

### 2020-08-08 ms/sections/discussion.tex
jonathans comments implemented

%\textbf{Jonathan's take: too much push in favor of selection, without really having much evidence. Simple models of genetic drift are ruled out and we don’t think it is plasticity. However, our primary test for adaptation—looking for parallel responses across islands—also was falsified. We suggest, however, that selection could be involved, and suggest several ways this might have happened.}

@@

\paragraph{A role of neutral drift is unlikely} Differences in organismal traits between environments are not necessarily proof of adaptation or selection, and genetic drift may result in patterns similar to local adaptation \citep{Miles2019}. Nevertheless, two lines of evidence from our data suggest that this scenario may be implausible. First, we found little evidence for a role of phenotypic isolation-by-distance (spatial autocorrelation) in explaining the differences we report. We did detect a significant signal of isolation-by-distance on Eleuthera, but there were no differences in dewlap coloration between habitats on this island. Second, we detected differences between habitats at relatively small spatial scales, most of the time between sites 5-10km apart, sometimes a few hundred meters away, on Bimini for example. Such small-scale differences would be unlikely under strong gene flow \citep{Richardson2014}. Our study islands lack geographic barriers to the movement of \textit{A. sagrei}, which have been shown to be highly mobile \citep{Kamath2018}, implying widespread gene flow across sites and habitats. Moreover, habitat-populations within each island were found to be non-monophyletic and often share identical haplotypes, based on phylogenetic analysis of mitochondrial DNA sequences (van de Schoot et al. unpublished thesis), suggesting gene flow between habitats may be widespread. Our results align with previous documented cases of persistent dewlap color divergence despite gene flow in multiple species of anoles, sometimes in relation to environmental conditions. \citet{Ng2012} and \citet{Ng2016} found divergent dewlap coloration in the face of gene flow between subspecies of \textit{A. distichus} across Hispaniola, and proposed this as a mechanism of reproductive isolation in the early stages of speciation \citep{Ng2011, Lambert2013, Ng2017}. \citet{Stapley2011} found that dewlap color polymorphism was maintained in the absence of genetic structure between populations of \textit{A. apletophallus} from Panama. \citet{Thorpe2002a} found that divergence in dewlap coloration matched habitat-type better than mitochondrial lineage in \textit{A. roquet} on Martinique, and a convergent pattern was found in \textit{A. trinitatis} on the featureless island of St Vincent \citep{Thorpe2002b}. Finally, regionally-distinct body coloration, but not dewlap coloration, is present in \textit{A. conspersus} on another small island, Grand Cayman, where no physical barriers to gene flow exist \citep{Macedonia2001}.\\

@@

\paragraph{Implications in the context of speciation} Local adaptation can be a precursor to ecological speciation, a process that may have given rise to the adaptive radiation of \textit{Anolis} lizards \citep{Harmon2003, Gavrilets2009}. Ecologically-mediated divergence of a sexual signal may be a potent path to the evolution of reproductive isolation through divergent sexual selection \citep{Reynolds2007, Servedio2011}. Evidence suggests that dewlap coloration could take this role in anoles \citep{Ng2011, Lambert2013, Geneva2015, Ng2017}, or at least that it is frequently involved in species recognition \citep{Williams1969, Williams1977, Losos1985, Macedonia1994, Fleishman2000, Macedonia2013, Ingram2016, Baeckens2018}. Although a correlation between dewlap coloration and reproductive isolation is not detected at the phylogenetic scale of the whole genus \citep{Nicholson2007, Harrison2012, Ingram2016}, sexual signals are often evolutionarily very labile \citep{Kraaijeveld2011}, and the anole dewlap in particular has been capable of rapid macroevolution throughout its phylogenetic history \citep{Nicholson2007}. For example, \textit{A. conspersus} on Grand Cayman evolved a UV-blue dewlap from an ancestral orange dewlap in 2 to 3 million years \citep{Macedonia2001}. We present evidence of multiple cases of within-island dewlap divergence over small geographical scales in \textit{A. sagrei} across the West Indies. While these intra-island populations do not appear to be in the process of speciation, our results suggest that the anoline dewlap has enough micro-scale, local adaptive potential to contribute to reproductive isolation, should it be recruited for assortative mating.

>>>>

%firef\textbf{Jonathan's take: too much push in favor of selection, without really having much evidence. Simple models of genetic drift are ruled out and we don’t think it is plasticity. However, our primary test for adaptation—looking for parallel responses across islands—also was falsified. We suggest, however, that selection could be involved, and suggest several ways this might have happened.}

@@

\paragraph{A role of neutral drift is unlikely} Differences in organismal traits between environments are not necessarily proof of adaptation or selection, and genetic drift may result in patterns similar to local adaptation \citep{Miles2019}. Nevertheless, two lines of evidence from our data suggest that this scenario may be implausible. First, we found little evidence for a role of phenotypic isolation-by-distance (spatial autocorrelation) in explaining the differences we report. We did detect a significant signal of isolation-by-distance on Eleuthera, but there were no differences in dewlap coloration between habitats on this island. Second, we detected differences between habitats at relatively small spatial scales, most of the time between sites 5-10km apart, sometimes a few hundred meters away, on Bimini for example. Such small-scale differences would be unlikely under strong gene flow \citep{Richardson2014}. Our study islands lack geographic barriers to the movement of \textit{A. sagrei}, which have been shown to be highly mobile \citep{Kamath2018}, implying widespread gene flow across sites and habitats. Moreover, habitat-populations within each island were found to be non-monophyletic and often share identical haplotypes, based on phylogenetic analysis of mitochondrial DNA sequences (van de Schoot et al. unpublished thesis), suggesting gene flow between habitats may be widespread. Our results align with previous documented cases of persistent dewlap color divergence despite gene flow in multiple species of anoles, sometimes in relation to environmental conditions. \citet{Ng2012} and \citet{Ng2016} found divergent dewlap coloration in the face of gene flow between subspecies of \textit{A. distichus} across Hispaniola, and proposed this as a mechanism of reproductive isolation in the early stages of speciation \citep{Ng2011, Lambert2013, Ng2017}. \citet{Stapley2011} found that dewlap color polymorphism was maintained in the absence of genetic structure between populations of \textit{A. apletophallus} from Panama. \citet{Thorpe2002a} found that divergence in dewlap coloration matched habitat-type better than mitochondrial lineage in \textit{A. roquet} on Martinique, and a convergent pattern was found in \textit{A. trinitatis} on the featureless island of St Vincent \citep{Thorpe2002b}. Finally, regionally-distinct body coloration, but not dewlap coloration, is present in \textit{A. conspersus} on another small island, Grand Cayman, where no physical barriers to gene flow exist \citep{Macedonia2001}.

@@

\paragraph{Implications in the context of speciation} Local adaptation can be a precursor to ecological speciation, a process that may have given rise to the adaptive radiation of \textit{Anolis} lizards \citep{Harmon2003, Gavrilets2009}. Ecologically-mediated divergence of a sexual signal may be a potent path to the evolution of reproductive isolation through divergent sexual selection \citep{Reynolds2007, Servedio2011}. Evidence suggests that dewlap coloration could take this role in anoles \citep{Ng2011, Lambert2013, Geneva2015, Ng2017}, or at least that it is frequently involved in species recognition \citep{Williams1969, Williams1977, Losos1985, Macedonia1994, Fleishman2000, Macedonia2013, Ingram2016, Baeckens2018}. Although a correlation between dewlap coloration and reproductive isolation is not detected at the phylogenetic scale of the whole genus \citep{Nicholson2007, Harrison2012, Ingram2016}, sexual signals are often evolutionarily very labile \citep{Kraaijeveld2011}, and the anole dewlap in particular has been capable of rapid macroevolution throughout its phylogenetic history \citep{Nicholson2007}. For example, \textit{A. conspersus} on Grand Cayman evolved a UV-blue dewlap from an ancestral orange dewlap in 2 to 3 million years \citep{Macedonia2001}. We present evidence of multiple cases of within-island dewlap color divergence over small geographical scales in \textit{A. sagrei} across the West Indies. While these intra-island populations do not appear to be in the process of speciation, our results suggest that the anoline dewlap could have enough micro-scale, local adaptive potential to contribute to reproductive isolation, should it be recruited for assortative mating.
